A comparison on R tools for nonlinear least squares modeling

### 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd

onls
crsnls
NISTnls
nlshelper
nlsic
nlsMicrobio
nlsmsn
nls.multstart
nlstac
nlstools
easynls
nlraa
optimx ??

It is likely that there are a number of other tools that could be relevant to
this comparison in the Bioconductor (??ref) collection and on repositories like
Github (??ref) and Gitlab (??ref), but we will not pursue those possibilities 
here.

@@

The tools for nonlinear estimation we consider have been built with different
goals and contexts. We will make our comparison in a number of directions.
We will make a number of specific notes, but our main commentary will consider
the following categories.

@@

@@

@@

>>>>

??put in proper refs to each

onls -- when there are errors in variables, this minimizes the orthogonal 
residuals

crsnls -- Controlled random search. No vignette!

NISTnls -- NIST problems for testing nls methods

nlshelper -- A few utilities for summarizing, testing, and plotting non-linear
    regression models fit with nls(), nlsList() or nlme()

nlsic -- We solve non linear least squares problems with optional
    equality and/or inequality constraints.

nlsMicrobio -- Data sets and nonlinear regression models dedicated to predictive microbiology.

nlsmsn -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) regression, details in Garay, Lachos and Abanto-Valle (2011)

nls.multstart -- Non-linear least squares regression with the Levenberg-Marquardt algorithm using multiple starting values for increasing the chance that the minimum found is the global minimum.

nlstac -- Set of functions implementing the algorithm described in Fernandez 
    Torvisco et al. (2018) for fitting separable nonlinear regression curves.
    ?? check ref and add to bib.??

nlstools -- Several tools for assessing the quality of fit of a gaussian nonlinear model are provided.

easynls -- Fit and plot some nonlinear models. ?? no vignette!

nlraa -- a set of nonlinear *selfStart* models. Most include analytic Jacobian
code. (@MiguezNLRAA2021)

optimx -- not really a nonlinear least squares method, but has general optimizers
that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
squares.

It is likely that there are a number of other tools that could be relevant to
this comparison in the Bioconductor (??ref) collection and on repositories like
Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
not pursue those possibilities here.

@@

The tools for nonlinear estimation we consider have been built with different
goals and contexts. We will make our comparison in a number of directions.
We will make a number of specific notes, but our main commentary will consider
the following categories. These overlap, for example in that the emphasis on modeling
versus solving the nonlinear least squares problem will result in different
output.f

@@

### Constraints on parameters

Reality often imposes conditions on parameters. While general constraints can
be difficult to include, bounds on parameters can be accommodated by many 
solvers. For some reason, fixed parameters (masks) are not often considered,
but are very convenient when a parameter has a commonly accepted value that we
later wish to allow to be estimated from the data.

@@

While the aim of nonlinear estimation is to return a set of parameters for a
particular model, there is a variety of additional information which may be
provided. From the perspective of the nonlinear least squares problem, there
are measures of work and information on the quality of the solution, while
for statistical purposes measures of uncertainty in the parameters are helpful.

Some tools allow constraints to be imposed on the solution, and it is helpful
to know if these constraints are active at the proposed solution.

@@

## An illustrative example

The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
task which is seemingly straightforward but turns out to be quite nasty. Because
this problem has a very succinct statement, it often provides the "short 
reproducible example" much beseeched of contributors to R mailing lists.

The data and a graph of it is given below.

```{r ex01, echo=TRUE}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
weeddf <- data.frame(tt, weed)
plot(weeddf, main="Hobbs weed infestation data")
```

Three suggested models for this data are (with names to allow for easy reference)

**Logistic3U**:
$$  y \approx  b_1 / (1 + b_2 * exp(- b_3 * t)) $$

**Logistic3S**:
$$ y \approx  100 * c_1 / (1 + 10 * c_2 * exp(- 0.1 * c_3 * t)) $$

**Logistic3T**:
$$ y \approx Asym / (1 + exp((xmid - t)/scal)) $$

The functions above are equivalent, but the first is generally more awkward
to solve numerically due to its poor scaling. The parameters of the three forms
are related as follows:

$$   Asym =  b_1 = 100 * c_1 $$
$$ exp(xmid/scal)  =  b_2 = 10 * c_2 $$
$$ 1/scal  =  b_3 $$
To allow for simpler discussion, let us say that the parameters for a vector
$p$ and the model function is called $model(p)$. The residuals can be written
either as

$$ r_1 = y - model(p)$$

or

$$ r = model(p) - y $$

since when squared the loss function has an identical value. In both cases, the
residual vector is a function of $p$. The second form
has a minor advantage in avoiding a potential sign error 
if we need to differentiate to get gradients etc.

### Loss function

The general approach of nonlinear optimizers, of which nonlinear least squares
solvers are usually a subset, is to try to reduce the loss function (or objective
function) defined by the sum of squares of the residuals ($y - model()$) by 
altering the model parameters from some starting guess. In all general 
methods of which we are aware, this is an iterative process.

We note several useful quantities

Loss function:
$$ S(p) = r' r$$
Gradient of $S(p)$:
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ \partial ^2 S(p) / {\partial p_i \partial p_j} $$
?? how to succinctly write the Hessian in terms of J' J

`nlsr` tries to find solutions from very poor initial parameter values. While
this is not always possible, `nslr` has generally been able to succeed in finding
solutions, even when all parameters are started at 1. Let us compare its results
with those of `nls()` and `nlsLM()`.

### Problem setup

```{r ex02set, echo=TRUE}
# model formulas
frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
#
# Starting parameter sets
stu1<-c(b1=1, b2=1, b3=1)
sts1<-c(c1=1, c2=1, c3=1)
stt1<-c(Asym=1, xmid=1, scal=1)
```

-------------------------------------------------
